% type of document
\documentclass{article}
% new section new page
\usepackage{titlesec}
\newcommand{\sectionbreak}{\clearpage}
% hyper-references
\usepackage{hyperref}
% align equation
\usepackage{amsmath}
% figures such as square
\usepackage{amssymb}
% appendix
\usepackage[toc,page]{appendix}
% title
\title{Personal notes on gravitational waves}
% author
\author{Jose Perdiguero Garate}

\begin{document}
\maketitle
\tableofcontents

\section*{Notation and conventions}

Throughout this notes I will be using the metric signature
\begin{equation}
    \left(-1, +1, +1, +1\right).
\end{equation}
Greek indices run over the four dimensional manifold spacetime, 
from $0$ to $3$, whereas latin indices run over only the spatial 
dimensions, from $1$ to $3$. Additionally, I am using the Einstein 
summation convention, where, repeated indices, indicate a summation.

The Minkowski metric tensor is written as $\eta_{\mu\nu}$ and 
an arbitrary metric tensor is $g_{\mu\nu}$. Partial derivatives 
are denoted by $\partial$ which can be acting on the four dimensions
or three dimensions, and can be written with a commas $g_{\mu\nu,\alpha}$. 
Covariant derivatives are defined with the symmetric standard connection
$\nabla$, which is compatible with the metric tensor such that 
$\nabla_{\alpha}g_{\beta\gamma} = g_{\beta\gamma;\alpha} = 0$
where I used semicolons for its denotation.

\section{Linear general relativity}

In this section I am presenting a general overview on how to build-up
the Einstein's field equation, given a background metric tensor and a 
perturbation tensor.

\subsection{General overview}

Decomposed the metric tensor as the sum of a background metric, in this case 
the flat spacetime Minkowski metric $\eta_{\mu\nu}$,plus a perturbation 
$h_{\mu\nu}$ as follows
\begin{equation}
    \label{metric tensor}
    g_{\mu\nu} =  \eta_{\mu\nu} + h_{\mu\nu}.
\end{equation}
Using the definition of the Kronecker delta object it is straightforward to
obtain the inverse tensor of the perturbation
\begin{equation}
    \label{definition delta}
    \delta^{\mu}_{\nu} = g^{\mu\sigma}g_{\sigma\nu}.
\end{equation}
Replacing Eq.\eqref{metric tensor} in to Eq.\eqref{definition delta}, and
neglecting second order terms in the perturbation field, leads to
\begin{equation}
    h^{\mu\nu} = -\eta^{\mu\alpha}\eta^{\nu\beta}h_{\alpha\beta},
\end{equation}
notice that, in order to upper/lower the indices of the perturbation I am only
using the background metric.\footnote{Including the perturbation tensor leads
to second order terms, which I am ignoring.} Schematically, second order terms
are neglected 
\begin{align}
    h_{\mu\nu}h_{\alpha\beta} & \sim 0, & 
    h_{\mu\nu}\partial_{\gamma} h_{\alpha\beta} & \sim 0, &
    \partial_{\delta} h_{\mu\nu}\partial_\gamma h_{\alpha\beta} & \sim 0.
\end{align}
This is the general overview of the fundamental field of general relativity,
which at its core is the metric tensor. In the following subsection, I will
be computing the Einstein's field equations for the metric tensor written 
in Eq.\eqref{metric tensor}.

\subsection{Geometrical objects}

The first object that is required to compute the Einstein's field equations is the
connection. Working on a torsion-free manifold, the Levi-Civita connection is written
as
\begin{equation}
    \label{levi civita connection}
    \Gamma^{\mu}{}_{\alpha\beta} = \frac{1}{2}g^{\mu\rho}\left(
        \partial_{\alpha}g_{\rho\beta} + \partial_{\beta}g_{\rho\alpha} -
        \partial_{\rho}g_{\alpha\beta}.
    \right)
\end{equation}
Replacing Eq.\eqref{metric tensor} in to Eq.\eqref{levi civita connection} leads to
\begin{equation}
    \Gamma^{\mu}{}_{\alpha\beta} = \frac{1}{2}\left(\eta^{\mu\rho} - h^{\mu\rho}\right)
        \left(\partial_{\alpha}\eta_{\rho\beta} 
        + \partial_{\beta}\eta_{\rho\alpha} 
        - \partial_{\rho}\eta_{\alpha\beta}
        + \partial_{\alpha}h_{\rho\beta} 
        + \partial_{\beta}h_{\rho\alpha} 
        - \partial_{\rho}h_{\alpha\beta}\right).
\end{equation}
Notice that, the only non-trivial contributions are the ones that are linear in the
perturbation, additionally, the partial derivatives of the Minkowski's metric tensor vanishes,
therefore, the connection coefficients are reduced to
\begin{equation}
    \label{levi civita connection flat + perturbation}
    \Gamma^{\mu}{}_{\alpha\beta} = \frac{1}{2}\eta^{\mu\rho}
        \left(\partial_{\alpha}h_{\rho\beta} + \partial_{\beta}h_{\rho\alpha} 
        - \partial_{\rho}h_{\alpha\beta}\right).
\end{equation}

Next, compute the Riemann curvature tensor
\begin{equation}
    \label{riemann curvature tensor}
    \mathcal{R}^{\rho}{}_{\sigma\mu\nu} = \partial_{\mu}\Gamma^{\rho}{}_{\sigma\nu}
    - \partial_{\nu}\Gamma^{\rho}{}_{\sigma\mu} 
    + \Gamma^{\gamma}{}_{\nu\sigma}\Gamma^{\rho}{}_{\mu\gamma}
    + \Gamma^{\gamma}{}_{\mu\sigma}\Gamma^{\rho}{}_{\nu\gamma},
\end{equation}
however, instead of computing directly from the above equation, it is convenient to 
notice the structure of the curvature tensor. The last two terms are quadratic in the 
Levi-Civita connection, and, since the connection is written with perturbation, then,
square terms in the connection vanishes, reducing the Riemann curvature tensor to
\begin{equation}
    \label{curvature tensor flat}
    \mathcal{R}^{\rho}{}_{\sigma\mu\nu} = \partial_{\mu}\Gamma^{\rho}{}_{\sigma\nu}
    - \partial_{\nu}\Gamma^{\rho}{}_{\sigma\mu}.
\end{equation}
Replacing Eq.\eqref{levi civita connection flat + perturbation} in to 
Eq.\eqref{curvature tensor flat} leads to
\begin{equation}
    \mathcal{R}^{\rho}{}_{\sigma\mu\nu} = \frac{1}{2}\eta^{\rho \alpha}\partial_{\mu}\left(
        \partial_{\sigma}h_{\alpha\nu} + \partial_{\nu}h_{\sigma\alpha}
        - \partial_{\alpha}h_{\sigma\nu}\right) +
        \frac{1}{2}\eta^{\rho \alpha}\partial_{\nu}\left(
        \partial_{\sigma}h_{\alpha\mu} + \partial_{\mu}h_{\sigma\alpha}
        - \partial_{\alpha}h_{\sigma\mu}\right),
\end{equation}
the above expression, can be simplified to
\begin{equation}
    \mathcal{R}^{\rho}{}_{\sigma\mu\nu} = \frac{1}{2}\eta^{\rho\alpha}\left(
        \partial_{\mu}\partial_{\sigma}h_{\alpha\nu} 
        - \partial_{\mu}\partial_{\alpha}h_{\sigma\nu}
        - \partial_{\nu}\partial_{\sigma}h_{\mu\alpha}
        + \partial_{\nu}\partial_{\alpha}h_{\mu\sigma}\right).
\end{equation}

From the Riemann tensor, it is straightforward to compute the Ricci tensor, by
contracting their respective indices
\begin{equation}
    \mathcal{R}_{\sigma\nu} = \mathcal{R}^{\mu}{}_{\sigma\mu\nu}.
\end{equation}
A direct computation shows the structure of the Ricci tensor
\begin{equation}
    \label{ricci tensor flat + perturbation}
    \mathcal{R}_{\sigma\nu} = \frac{1}{2}\left(
        \partial^{\alpha}\partial_{\sigma}h_{\alpha\nu} 
        -  \square h_{\sigma\nu} - \partial_{\nu}\partial_{\sigma}h
        + \partial_{\nu}\partial^{\alpha}h_{\alpha\sigma}\right),
\end{equation}
where $\square$ is the d'Alembert operator and $h$ is the trace of the perturbation.

In the same spirit, the curvature scalar can be obtained directly through
the contraction of the Ricci tensor
\begin{equation}
    \mathcal{R} = g^{\mu\sigma}\mathcal{R}_{\mu\sigma}.
\end{equation}
This computation is straightforward
\begin{equation}
    \mathcal{R} = \left(\eta^{\mu\sigma} - h^{\mu\sigma}\right) \frac{1}{2}\left(
        \partial^{\alpha}\partial_{\sigma}h_{\alpha\nu} 
        -  \square h_{\sigma\nu} - \partial_{\nu}\partial_{\sigma}h
        + \partial_{\nu}\partial^{\alpha}h_{\alpha\sigma}\right).
\end{equation}
Neglecting second order terms in the perturbation field, the scalar curvature is given
by
\begin{equation}
    \label{curvature scalar flat + perturbation}
    \mathcal{R} = \partial_{\mu}\partial_{\sigma}h^{\mu\sigma} - \square h
\end{equation}

Now, we can compute the Einstein's field equations without a cosmological constant
\begin{equation}
    \label{Einsteins field equation}
    \mathcal{R}_{\mu\nu} - \frac{1}{2}\mathcal{R}g_{\mu\nu} = \frac{8\pi G}{c^4}
    \mathcal{T}_{\mu\nu}
\end{equation}
where $\mathcal{T}_{\mu\nu}$ is the energy momentum tensor. Replacing 
Eq.\eqref{ricci tensor flat + perturbation} and Eq.\eqref{curvature scalar flat + perturbation}
in to Eq.\eqref{Einsteins field equation} leads to
\begin{equation}
        \partial^{\alpha}\partial_{\mu}h_{\alpha\nu} 
        -  \square h_{\mu\nu} - \partial_{\nu}\partial_{\mu}h
        + \partial_{\nu}\partial^{\alpha}h_{\alpha\mu}
        - \left(\partial_{\alpha}\partial_{\beta}h^{\alpha\beta} - \square h\right)
        \left(\eta_{\mu\nu} + h_{\mu\nu}\right) = \frac{16\pi G}{c^4}\mathcal{T}_{\mu\nu}.
\end{equation}
Just like before, neglecting second order terms in the perturbation, the above equation
can be reduced to
\begin{equation}
    \label{einsteins equation flat + perturbation}
        \partial^{\alpha}\partial_{\mu}h_{\alpha\nu} 
        -  \square h_{\mu\nu} - \partial_{\nu}\partial_{\mu}h
        + \partial_{\nu}\partial^{\alpha}h_{\alpha\mu}
        - \eta_{\mu\nu}\partial_{\alpha}\partial_{\beta}h^{\alpha\beta} 
        - \eta_{\mu\nu}\square h = \frac{16\pi G}{c^4}\mathcal{T}_{\mu\nu}.
\end{equation}
The above equation, can be written in a much more compact manner by using the following
variable change\footnote{In the standard literature $X_{\mu\nu}$ is written as 
$\bar{h}_{\mu\nu}$, but I strongly believed this leads to confusions.}
\begin{equation}
    X_{\mu\nu} = h_{\mu\nu} - \frac{1}{2}\eta_{\mu\nu}h,
\end{equation}
which can be inverted through standard methods
\begin{equation}
    \label{variable change}
    h_{\mu\nu} = X_{\mu\nu} - \frac{1}{2}\eta_{\mu\nu}X
\end{equation}
where $X$ is the trace of the tensor $X_{\mu\nu}$, and the tensor $X_{\mu\nu}$, also satisfies
the relation $X = -h$. Replacing the variable change written in Eq.\eqref{variable change} in 
Eq.\eqref{einsteins equation flat + perturbation} and simplifying terms, leads to
\begin{equation}
    \label{efe X}
    \partial^{\alpha}\partial_{\mu}X_{\alpha\nu} + \partial_{\nu}\partial^{\alpha}X_{\alpha\mu}
    - \square X_{\mu\nu} - \eta_{\mu\nu}\partial_{\alpha}\partial_{\beta}X^{\alpha\beta}
    =  \frac{16\Pi G}{c^4}\mathcal{T}_{\mu\nu},
\end{equation}
which, in some sense has a more simple structure that Eq.\eqref{einsteins equation flat + perturbation},
and also contains the wave operator. Nonetheless, this does not look like a gravitational 
wave equations. In the next subsection, I will show you, how can you derived the gravitational
wave equation from the above expression using a gauge transformation.

\subsection{Gauge transformation}

Consider the infinitesimal gauge coordinate transformation
\begin{equation}
    \label{gauge transformation}
    x^{\mu} \longrightarrow x'^{\mu} =  x^{\mu} + \xi^{\mu},
\end{equation}
where $\xi^{\mu}$ is a small vector. Then, it is possible to obtain the relation
between and the inverse relation of the coordinate transformation of their respective
derivatives 
\begin{align}
    \label{gauge transformation derivatives}
    \frac{\partial x'^{\alpha}}{\partial x^{\beta}} & = \delta^{\alpha}_{\beta} 
    + \partial_{\beta} \xi^{\alpha} & 
    \frac{\partial x^{\alpha}}{\partial x'^{\beta}} & = \delta^{\alpha}_{\beta} 
    - \partial_{\beta} \xi^{\alpha}.
\end{align}

Using the above information, the metric tensor under a gauge coordinate transformation
changes as 
\begin{equation}
    g'_{\alpha\beta} = \frac{\partial x^{\mu}}{\partial x'^{\alpha}}
    \frac{\partial x^{\nu}}{\partial x'^{\beta}}g_{\mu\nu}.
\end{equation}
Using Eq.\eqref{gauge transformation derivatives} in the above equation, and neglecting
second order terms in the perturbation, leads to
\begin{equation}
    g'_{\alpha\beta} = g_{\alpha\beta} - g_{\alpha\nu}\partial_{\beta}\xi^{\nu}
    - g_{\mu\beta}\partial_{\alpha}\xi^{\mu}
\end{equation}
replacing the expressions for the metric tensor, see Eq.\eqref{metric tensor}, leads to
\begin{equation}
    \label{gauge transformation of perturbation}
    h'_{\alpha\beta} = h_{\alpha\beta} - \partial_{\beta}\xi_{\alpha} 
    - \partial_{\alpha}\xi_{\beta}.
\end{equation}
Is worth mention that the Riemann tensor is invariant under a gauge transformation. 
Therefore, we have the freedom to choose or fix the vector $\xi^{\mu}$ as we liked. 
Additionally, Eq.\eqref{gauge transformation of perturbation} is only valid using a 
Minkowski background, if we were working on a curved spacetime background, there will
be additional terms, the most general expression can be found in at the end of this 
section. As the rule of gauge transformation is written in 
Eq.\eqref{gauge transformation of perturbation}, then is trivial to compute the gauge
transformation of the auxiliary variable $X_{\mu\nu}$
\begin{equation}
    X_{\mu\nu} \to X'_{\mu\nu} = h'_{\mu\nu} 
    - \frac{1}{2}\eta_{\mu\nu}\eta^{\alpha\beta}h'_{\alpha\beta},
\end{equation}
replacing the transformation rule leads to
\begin{equation}
    X'_{\mu\nu} = h_{\mu\nu} - \partial_{\nu}\xi_{\mu} 
    - \partial_{\mu}\xi_{\nu}
    - \frac{1}{2}\eta_{\mu\nu}\eta^{\alpha\beta}\left(
    h_{\alpha\beta} - \partial_{\beta}\xi_{\alpha} - \partial_{\alpha}\xi_{\beta}\right).
\end{equation}
The above expression can be simplified 
\begin{equation}
    \label{gauge transformation X}
    X'_{\mu\nu} = X_{\mu\nu} - \partial_{\mu}\xi_{\nu} 
    - \partial_{\nu}\xi_{\mu} + \eta_{\mu\nu}\partial^{\sigma}\xi_{\sigma},
\end{equation}
where I used the definition of the $X_{\mu\nu}$ tensor written in Eq.\eqref{variable change}. 
Now, it is convenient to work with upper index, thus, using the inverse Minkowski metric we 
can upper every index
\begin{equation}
    X'^{\alpha\beta} = X^{\alpha\beta} - \eta^{\mu\alpha}\partial_{\mu}\xi^{\beta} 
    - \eta^{\nu\beta}\partial_{\nu}\xi^{\alpha} 
    + \eta^{\alpha\beta}\partial^{\sigma}\xi_{\sigma}.
\end{equation}
At this point we take the divergence of $X_{\mu\nu}$
\begin{equation}
    \partial_{\beta}X'^{\alpha\beta} = \partial_{\beta}X^{\alpha\beta}
    - \eta^{\mu\alpha}\partial_{\beta}\partial_{\mu}\xi^{\beta} 
    - \eta^{\nu\beta}\partial_{\beta}\partial_{\nu}\xi^{\alpha} 
    + \eta^{\alpha\beta}\partial_{\beta}\partial^{\sigma}\xi_{\sigma},
\end{equation}
after simplification of terms, leads to
\begin{equation}
    \partial_{\beta}X'^{\alpha\beta} = \partial_{\beta}X^{\alpha\beta}
    - \square \xi^{\alpha}.
\end{equation}
At this point, recall that we still have the freedom to choose the $\xi^{\mu}$. Therefore,
fixing 
\begin{equation}
    \label{gauge condition}
    \square \xi^{\alpha} = \partial_{\beta}X^{\alpha\beta} 
    \rightarrow \partial_{\beta}X'^{\alpha\beta} = 0,
\end{equation}
this is known as the Lorentz gauge, which vanishes the vast majorities of 
terms of Eq.\eqref{efe X}. The only, non-trivial term comes from the wave 
operator, leading to
\begin{equation}
    \square X_{\mu\nu} = -\frac{16\Pi G}{c^4}\mathcal{T}_{\mu\nu},
\end{equation}
which for the special vacuum case $\mathcal{T}_{\mu\nu} = 0$, reduces the 
above equation to 
\begin{equation}
    \square X_{\mu\nu} = 0.
\end{equation}
The above result, is known as the gravitational wave equation. In the next subsection we will
be dealing with the problem of counting the degrees of freedom of a gravitational wave.

\subsection{Degrees of freedom}

The perturbation has 16 independent components which must be determined. 
However, because it is defined as a symmetric tensor it only has 10 
independent components. Moreover, due to the gauge condition 
Eq.\eqref{gauge condition} there are four additional equations/constraint 
that must be satisfied, leaving only 6 independent components. Nonetheless, 
this choice, does not completely fixes the gauge, we still have the freedom 
to choose to fix the four components of the displacement vector $\xi^{\mu}$, 
and hence, reduced even further the number of independent components of the 
perturbation to 2 independent components. This is known as the residual gauge.

An intuitive way of seen this reduction of the independent components of the
perturbation tensor comes from knowing that initially there are 10 independent
components of a symmetric tensor of rank $2$ tensor. Then by using the Lorentz
gauge restriction 
\begin{equation}
    \partial_{\beta}X'^{\alpha\beta} = 0,
\end{equation}
which lead to four different equations, meaning that, there are four more 
constraint to impose, reducing the number of 10 independent components to
6. Then, we can perform another gauge transformation by an infinitesimal
vector displacement $\xi^{\mu}$ 
\begin{equation}
    X'_{\mu\nu} \rightarrow X'_{\mu\nu} + \xi_{\mu\nu},
\end{equation}
where $\xi_{\mu\nu}$ is defined as 
\begin{equation}
    \xi_{\mu\nu} \equiv \eta_{\mu\nu}\partial^{\alpha}\xi_{\alpha} 
    -\partial_{\mu}\xi_{\nu} -\partial_{\nu}\xi_{\mu}.
\end{equation}
Therefore, choosing properly the vector $\xi_{\mu}$ such that $\square \xi_{\mu\nu} = 0$,
it is possible to reduced 4 more degrees of freedom of the perturbation 
tensor. Leaving only $2$ independent components, which are known as the $2$
polarization, $+$ and $x$.

Exploiting the residual gauge symmetry, it is possible to eliminate
directly components of the perturbation tensor. Using Eq.\eqref{gauge transformation X}
it is possible to vanishes the trace and the spatial-temporal components of the 
perturbation $X = 0$ and $X_{0i} = 0$. As a consequence of this, the perturbation
and the variable change are the same $X_{\mu\nu} = h_{\mu\nu}$. From the Lorentz gauge
condition Eq.\eqref{gauge condition}
\begin{equation}
    \dot{h}_{00} + \partial_{i}h_{i0} = 0,
\end{equation}
from which, we infer that the temporal-temporal components is a function of only
the spatial coordinates. As this component does not depend on the time coordinate, (
which is our concern, because gravitational waves are time-dependent), we can set 
$h_{00} = 0$. In a nutshell, the constraint are
\begin{align}
    h_{\mu 0 } & = 0, & h & = 0, & \partial_{i}h_{ij} & = 0.
\end{align}
This is known as the \textit{transverse-traceless} gauge (TT).


\subsection{Scalar-Vector-Tensor decomposition}

Although we have found so far the gravitational wave equation coupled with a
energy-momentum tensor, in principle we are done. Now, the next task should be
to compute the Einstein's field equation for a given symmetry of the metric tensor.
However, we can take advantage of the \textit{scalar-vector-tensor} decomposition
to write down the independent components of Einstein's equations in a much more 
simple form.

First, lets decomposed the perturbation tensor as follows
\begin{align}
    h_{00} & = -2\phi, \\
    h_{0i} & = \partial_i B + S_i, \\
    h_{ji} & = h_{ij} + \partial_{i} F_{j} + \partial_{j} F_{i} - 2\psi\delta_{ij}, 
    + \left(\partial_{i}\partial_{j} - \frac{1}{3}\delta_{ij}\nabla^{2}\right)E,
\end{align}
where $\phi$, $B$, $\psi$ and $E$ are scalars functions, $S_i$ and $F_i$ 
are vectors whose divergence vanishes $\partial_{i}S_{i} = \partial_{i}F_{i} = 0$,
and $h_{ij}$ is a tensor whose divergence is trivial $\partial_{i}h_{ij} = 0$ and also
by construction must be traceless $h = 0$. Notice that, in principle, the perturbation
tensor has $10$ independent components, and from the decomposition, we have $4$ scalars,
$6$ components for the two vectors $-2$ constraints, and from the tensorial part
there are only $6$ independent components $-4$ constraints. Therefore, the decomposition
leads to $4 + \left(6 - 2\right) + \left(6 - 4\right) = 10$, which, of course, matches
the number of independent components of the perturbation. Details on how to derived this
decomposition can be found in Ref.\cite{poisson2014gravity} in section $5.5$.

In the same way, the vector displacement $\xi_{\mu}$ can also be decomposed
\begin{equation}
    \xi_{\mu} = \left(d_0, \partial_i d + d_i\right),
\end{equation}
and using the gauge transformation rule written in Eq.\eqref{gauge transformation X},
it is straightforward to derived how the components of the perturbation tensor transform
under a coordinate gauge transformation. Take for example the $h_{tt}$ component
\begin{equation}
    h'_{tt} = h_{tt} - 2\partial_{t}\xi_{t},
\end{equation}
which, replacing the values leads to
\begin{equation}
    2\phi' = 2\phi + 2\dot{d}_0,
\end{equation}
then, the rule of transformation can be written as
\begin{equation}
    \phi' = \phi + \dot{d}_0.
\end{equation}
Applying the same idea to the other scalars, vectors and tensor components of the
tensor perturbation, we find that 
\begin{align}
    \phi' & = \phi + \dot{d}_0, & B' & = B - d_0 - \dot{d}, & E & = E - 2d, \\
    S_i' & = S_i - \dot{d}_i, & F_i & = F_i - d_i.
\end{align}
Once the rules of gauge transformation are known for the STV decomposition, it is
straightforward to build-up gauge invariants using the components of the perturbation, 
these are gauge invariants are
\begin{align}
    \Phi & \equiv \phi + \dot{B} - \frac{1}{2}\ddot{E}, &
    \Theta & \equiv -2\psi - \frac{1}{3}\nabla^2E, &
    \Sigma_{i} & \equiv S_i - \dot{F}_i, &
    h_{ij} & \equiv h_{ij},
\end{align}
there are two scalar gauge invariants, one vector invariant and one tensor invariant
quantity.

A similar procedure can be applied to the stress energy momentum tensor
\begin{align}
    \label{energy momentum tensor svt}
    \mathcal{T}_{00} & = \rho, \\ 
    \mathcal{T}_{0i} & = \partial_{i}u + u_i, \\
    \mathcal{T}_{ij} & = \Pi_{ij} + \partial_{i}v_{j} + \partial_{j}v_{i} + p\delta_{ij}
    + \left(\partial_{i}\partial_{j} - \frac{1}{3}\delta_{ij}\nabla^{2}\right)\sigma,
\end{align}
where $\rho$, $u$, $p$ and $\sigma$ are scalar functions, $u_i$ and $v_i$ are vectors
whose divergence vanishes $\partial_{i}u_{i} = 0 $ and $\partial_{i}v_{i} = 0$, and $\Pi_{ij}$
is a traceless tensor (by construction), therefore $\Pi = 0$, whose divergence vanishes 
$\partial_{i}\Pi_{ij} = 0$. The natural next step, would be to build-up the rules of
gauge transformation of the components of the energy-momentum tensor, however because 
the energy-momentum tensor must be zero in the background, all the components
(perturbed) are gauge invariants, this is known as the Steward Walker lemma 
\cite{stewart1993advanced}.

From the conservation law, 
\begin{equation}
    \nabla_{\mu}\mathcal{T}^{\mu\nu} = \partial_{\mu}\mathcal{T}^{\mu\nu}
    + \Gamma^{\mu}{}_{\mu\lambda}\mathcal{T}^{\lambda\nu}
    + \Gamma^{\nu}{}_{\mu\lambda}\mathcal{T}^{\mu\lambda} \sim \partial_{\mu}\mathcal{T}^{\mu\nu}.
\end{equation}
Replacing Eq.\eqref{energy momentum tensor svt} into the above equation leads to the following
constraints
\begin{align}
    \nabla^{2}u & = \dot{\rho}, & 
    \nabla^{2}\sigma & = \frac{3}{2}\left(\dot{u} - p\right), & 
    \nabla^{2}v_{i} & = \dot{u}_{i}, & 
\end{align}

The next step, is to put all this ingredients together into Einstein's field equation. A 
straightforward computation leads to the following field equations
\begin{align}
    \label{efe linear with matter}
    \nabla^2 \Theta & = -\rho, \\
    \nabla^2 \Phi & = \left(\rho + 3p - 3\dot{u}\right), \\
    \nabla^2 \Sigma_{i} & = - 2S_{i}, \\
    \square h_{ij} & = -2\Pi_{ij}.
\end{align}
These are the set of differential equations, where there is only one wave equations coming
from the $h_{ij}$. The other fields obey a Poisson-like equation. Notice that, both sides of 
the Eqs.\eqref{efe linear with matter} are gauge invariants, as they should be.

\subsection{Gravitational waves in a curved spacetime}

Now, lets move on one step a head and use the FRW metric as the background metric
\begin{equation}
    g_{\mu\nu} = \bar{g}_{\mu\nu} + h_{\mu\nu},
\end{equation}
where the bar quantities make references to background objects. The distance between two
points can be compute using
\begin{align}
    \mathrm{d}s^{2} = a^{2}(\eta)\left(-\mathrm{d}\eta^2 
    + \delta_{ij}\mathrm{d}x^{i}\mathrm{d}x^{j}
    + h_{ij}\mathrm{d}x^{i}\mathrm{d}x^{j}\right),
\end{align}
where $h_{ij}$ is in the TT-gauge, and I am also working in conformal time, defined 
by the relation 
\begin{equation}
    \label{conformal time relation}
    a\mathrm{d}\eta = \mathrm{d}t.
\end{equation} 
Next, you can compute the Christoffel symbols, the Riemann curvature tensor and the 
Ricci scalar. These are the geometrical objects that you need to compute the Einstein's 
field equations. The perturbed Einstein's field equation leads to just one equation
\begin{equation}
    h''_{ij} + 2\mathcal{H}h'_{ij} - \nabla^{2}h_{ij} = 8\pi G a^{2}\bar{p}\Pi_{ij},
\end{equation}
where the time derivatives are take with respect to the conformal time, and the Hubble function
$\mathcal{H}$ is defined also in conformal time. Using Eq.\eqref{conformal time relation} you can
work the above equation into physics time
\begin{equation}
    \ddot{h}_{ij} + 3H\dot{h}_{ij} - \frac{\nabla^{2}}{a^{2}}h_{ij} = 8\pi G \bar{p}\Pi_{ij}.
\end{equation}
From the above expression it is not obvious the classification type of solutions. However,
by using the following variable change
\begin{equation}
    H_{ij}\left(\bold{k},\eta\right) = a h_{ij}\left(\bold{k},\eta\right),
\end{equation}
the differential equation turns to 
\begin{equation}
    H_{ij}\left(\bold{k},\eta\right) + \left(k^2 - \frac{a''}{a}\right)H_{ij}\left(\bold{k},\eta\right)
    = 16\pi G a^3\Pi_{ij}.
\end{equation}
Looking for analytical solutions in vacuum $\Pi_{ij} = 0$, there is a clear distinction between
two branches of solutions. The

To finalized this section, I just want to deduced the most general gauge transformation
of a rank 2 tensor, having a curved background. This is a more general expression that the one
found in Eq.\eqref{gauge transformation of perturbation}. 

\section{Chern-Simons gravity}

Consider the Chern-Simons action coupled with the Einstein-Hilbert action
\begin{equation}
    S = \int \mathrm{d}^4x 
\end{equation}

\section{Polynomial parity breaking}

The most general 


\bibliographystyle{utphys}
\bibliography{references}

\end{document}